---
title: "Class 12"
author: "Luigi Franco"
format: pdf
editor: visual
---

# Section 1 - 4

```{r}
#Reading the csv files
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
#checking the first couple data points in each 
head(counts)
head(metadata)
```

**Q1. How many genes are in this dataset?**

> There are 38694 genes in this dataset.

**Q2. How many ‘control’ cell lines do we have?**

> There are 4 'control' cell lines

```{r}
dim(counts)
```

```{r}
#Calculating the mean counts per gene of the control samples 
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

**Q3. How would you make the above code in either approach more robust? Is there a function that could help here?**

> I would switch the code part "/4" because if I change the number of sample the output will not be correct. Instead, I would put the code "ncol()" so that it always follow the precise numberof columns in the data frame used.

**Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)**

```{r}
#Question 4
treated <- metadata[metadata[,"dex"]=="treated",]
treated.counts <- counts[, treated$id]
treated.mean <- rowSums(treated.counts) / ncol(treated.counts)
head(treated.mean)

#Combining mean datas 
meancounts <- data.frame(control.mean, treated.mean)
colSums(meancounts)
```

```{r}
# Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.
plot(meancounts[,1],meancounts[,2], xlab="Control", ylab="Treated")

#Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom_?() function would you use for this plot?
library(ggplot2)
ggplot(meancounts, aes(x=meancounts[,1], y=meancounts[,2])) +
  geom_point() +
  xlab("Control") +
  ylab("Treated")

```

**Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom\_?() function would you use for this plot?**

> geom_point()

**Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this?**

> log = "xy"

```{r}

plot(meancounts[,1], meancounts[,2], log="xy",
     xlab="log Control", ylab="log Treated")
```

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)

#getting rid of data that have zero expression
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

**Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?**

> The "arr.ind = TRUE" argument makes which() return row and column indices instead of the default index number, which is useful in knowing where the zeros are. We take the first column because we only care about the row indices (genes) to remove, not which column the zero appeared in. The function unique() is necessary becuase the same row could appear twice if a gene has zeros in both columns, so it just makes sure we have a clean list to avoid errors.

```{r}
#filtering data to see which genes are up or down-regulated
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

**Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level?**

> 250

**Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level?**

> 367

```{r}
#Questions 8 & 9
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)

paste("Up:", sum(up.ind))
paste("Down:", sum(down.ind))
```

**Q10. Do you trust these results? Why or why not?**

> I don't trust these results because we have not yet taken into account their p-values for statistically significant data. So these results cold be misleading.

# Section 5

```{r}
#Creating the dds object (DeSeqDataSet)
library(DESeq2)
citation("DESeq2")
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

# Section 6

```{r}
#Plotting PCA for analysis
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = c("dex"))
```

```{r}
#Building the PCa pot from scratch using ggplot
pcaData <- plotPCA(vsd, intgroup=c("dex"), returnData=TRUE)
head(pcaData)
#Calculating percent variacne
percentVar <- round(100 * attr(pcaData, "percentVar"))
#Plotting w/ ggplot
ggplot(pcaData) +
  aes(x = PC1, y = PC2, color = dex) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```

# Section 7

```{r}
#Checking results - error happens
## results(dds)
#creating dds from DESeq
dds <- DESeq(dds)
#Calling in the results 
res <- results(dds)
res

#using the summary function to summarize
summary(res)

#changing the alpha value from default 0.1 to 0.05
res05 <- results(dds, alpha=0.05)
summary(res05)
```

# Section 8

```{r}
#Checking new package 
library("AnnotationDbi")
library("org.Hs.eg.db")
columns(org.Hs.eg.db)
```

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
head(res)
```

**Q11. Run the mapIds() function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called res**$entrez, res$uniprot and res\$genename.

```{r}
#Question 11
res$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")

res$uniprot <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="UNIPROT",
                     keytype="ENSEMBL",
                     multiVals="first")

res$genename <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="GENENAME",
                     keytype="ENSEMBL",
                     multiVals="first")

head(res)
```

```{r}
#Arrangin and viewing the results (Q11) with adjusted p-value
ord <- order( res$padj )
head(res[ord,])

#writing out orderly significant results w/ annotations
write.csv(res[ord,], "deseq_results.csv")
```

# Section 9

```{r}
#Creating a Volvano Plot
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")

#making some adjustment to improve the plot, I decided to add red lines
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="red", lty=2)
abline(h=-log(0.05), col="red", lty=2)
```

```{r}
#changing the color of the points. I decided the colors I liked best
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "darkgreen" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="red", lty=2)
abline(h=-log(0.1), col="red", lty=2)
```

```{r}
#Further improving the customization
library(EnhancedVolcano)
x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')

#this looks pretty good! :)
```
